Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2019-08-11 Response_to_Reviewers.tex

\bigskip

We want to thank the two reviewers for their thorough evaluation and discussion of our paper, as well as the challenging questions they asked. We have revised the manuscript in a way that addresses most of the comments. In this response, we first want to address the two main questions that arose for both reviews. The first one (A) relates to the original mis-specification/ambiguities between the social and spatial components of the modelling process. The second one (B) relates to the contribution of the approach in terms of potential use for model makers, for domain experts and in operational contexts.

 \medskip
A. In terms of the ambiguity between the social and spatial component 
\begin{itemize}
\item A.1. We have precised and clarified the argument of the paper (mainly in introduction) by qualifying our contribution, which is to provide an operational framework for studying the influence of geometric structures of space on the results of a geosimulation model. We have thus distinguished two effects of geographical space in simulation modelling: the ones that are part of the model, reflecting for example the role played by distance on decreasing agents' perception or potential social interactions; and the ones which act as geometric structures "on which" the social elements simulated interact, i.e. geometric constraints of the simulation. The first type of effects is addressed by traditional modelling. Our original contribution is to tackle the second type. %\comment[CC]{done in 1.5 and 1.8}

\item A.2. We have clarified the terms used in our paper introduction to distinguish between: \textbf{sociospatial systems} (i.e. groups of social agents whose behaviour is constrained by their position in geographical space. This term refers to empirical systems, which are the targets of the modelling process), \textbf{interactions} (what happens between agents and between agents and their environment, in the model as well as in the empirical system) and \textbf{geometric structure}, a term we reserve for qualifying the geometric outputs of the spatial generator which are used as initial spatial conditions in the model. When we talk about the generator we actually used in the paper, we call it a "density grid generator" because this is the only geometric structure it generates, whereas a "spatial generator" could potentially generate graph or point structures. %\comment[CC]{done in 1.5}

@@

 \item A.3. We have modified figure 1 to express the clarified approach visually. Naming the generator parameter "gamma" has helped reduced the confusion that the term 'meta-parameter' was inducing. We have also replaced the expressions "quantitative results" by "sensitivity analysis" and "qualitative results" by "variations by types of geometric structure", in the figure as in the subsections of the paper. %\comment[CC]{done}
\end{itemize}

 \medskip
B. To highlight limitations as well as promising perspectives of improvement, answer the concern about the genericity of the method and expose its links with the conceptual and empirical domains:

 \begin{itemize}

  \item B.1. We have added a discussion of how segregation indicators can themselves be influenced by geometric structures (because they give the same weight to cells of different density). This is a limitation of the method as it is but also an avenue for promoting stronger indicators in empirical research. % \comment[MLT]{j'ai formulé un paragraphe court sur ce point en discussion, en lien avec le point B.3}

  \item B.2. To inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, We added a paragraph in discussion, explaining how our approach allows to go back to questioning empirical cases from the model and questioning our concepts (for example how 'neighbourhood' is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences). \comment{il faut reprendre ces elements dans l'article}

  \item B.3. As a perspective to reduce this bias, we could compute ego-neighbourhoods rather than a regular grid (another paper). We could also in another paper vary the size of grid and its cell shape (cf. MAUP), but also generate other types of geometric structures to feed into geosimulation models (for example points or networks). % \comment{j'ai formulé un paragraphe court sur ce point en discussion, en lien avec le point B.1}
\end{itemize}

@@

\bigskip

%\textbf{to do list:}

 % \textbf{Sociospatial system} pour parler des systèmes empiriques, \textbf{interactions} pour parler de ce qui se passe dans le modele,  \textbf{geometric structures} pour parler des outputs issus du générateur et  parler de \textbf{density grid generator} quand on développe le nôtre. \comment{done}

 % caser" Jessop (2008) points in particular to the interdependence between social phenomena and the spatial environment where they take place. We model the social system as the interaction dynamics of agents and spatial units. The spatial generator produces the geometric support for these interactions." in intro 1.4 \\ \comment[MLT]{To avoid redundancies, I have simply added the reference to Jessop when we describe Figure 1.}

@@

  \item it as not explicit enough and we revised the presentation of our approach (which is generic and not empirical) in intro (especially fig. 1). We thank reviewer 2 for highlighting this limitation.

  \medskip

  \textit{Furthermore, there are now examples of spatial models that go beyond administrative neighborhoods and overimposed spatial boundaries, i.e., egohoods (e.g., https://onlinelibrary.wiley.com/doi/abs/10.1111/1745-9125.12006), which could be interesting to see implemented in ABM.}

  $\rightarrow$ we added a paragraph on the limits of our density grid approach in discussion, in which we mention the possibility to use egohoods to better measure segregation.

### 2019-08-11 paper.tex
Update on Overleaf.

\author[5]{Marion Le Texier}
\affil[5]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}

@@

\begin{abstract}
Although simulation models of socio-spatial systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually not modified systematically in socio-spatial models, thus leaving unexplored the effect of initial spatial arrangements on the interactions of agents with one another as well as with their environment. In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic spatial configuration generator in order to create density grids with which socio-spatial simulation models are initialised. We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models. Furthermore, this effect is sometimes larger than the effect of parameters' value change.
\end{abstract}

@@

Self-organization has been shown to be a central feature of socio-spatial systems in general and of cities in particular ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}. In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}. Complexity is partially due to bifurcations, which are determinant in socio-spatial systems~\citep{Wilson1981, Wilson2002}. Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures. In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as any other initial conditions, but of more interest to the geographer. \\

@@

In this paper, we aim to provide an operational framework for studying the influence of geometric structures of space on the results of a simulation model representing a sociospatial system. {\bf Sociospatial systems} are understood here as groups of social agents whose behaviour is constrained by their position in geographical space, in the empirical world. The aim of the geosimulation models we are interested in is to represent, simulate and explain the dynamics of these systems. In the sociospatial systems as well as in the models representing them, agents are in {\bf interaction} among themselves and with their geographical environment. The {\bf geometric structure} of the environment defined at the initialisation of the model and its influence on the model outputs is our object in this paper. We present how to generate the geometric outputs to are used as initial spatial conditions in the model. To this end, we use the example of a density grid generator and feed its outputs to two application models (Schelling and Sugarscape). In no way do we pretend to provide a full exploration of these two particular models, of their attractors and/or potential policy implications. Instead, we present a way of performing a sensitivity analysis to initial spatial conditions of models generated systematically. The generator being controlled by its own parameters, we can then relate the parameters used to generate initial spatial conditions to the variation simulation outcomes. The purpose is two-fold: (i) to test the robustness of simulation results to small variations of generator parameters and (ii) to study the non-trivial effects of typical categories of geometric structures (monocentric \textit{vs.} polycentric for example) on the results of a given model. Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies to our knowledge. In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.

@@

The spatial generator applies an urban morphogenesis model \citep{Batty2007} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}. An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily into our workflow. Grids are generated through an iterative process which, starting from an empty grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to a patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (i.e. its 8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $t_f$ is reached. The grid then has a population of $t_f \cdot N$ (the population lost due to diffusion process to the outside is reallocated through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four generator parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment. We furthermore fix the total population at $t_f\cdot N = 100,000$, and determine therein the number of steps needed at a given $N$. Typical value ranges for the  parameters will be taken as, following \citet{raimbault2018calibration}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$. We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configurations that can be generated.

@@

The implementations of the models were done from NetLogo. We modified the sugarscape version with wealth of NetLogo model library (to be able to explore it intensively) and we implemented from scratch the Schelling model. Both pseudo-codes are available in~\ref{app:code}, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}. Density grids are also available at this address. Simulation data are available for reproducibility on the dataverse repository at \url{https://doi.org/10.7910/DVN/A0N5GV}.

@@

We measure the distance of the phase diagrams for all density grids with respect to the reference phase diagram computed on the default initial spatial condition setup (a bi-centric symmetrical non toroidal configuration) using the measure defined in equation~\ref{eq:phase-distance}. For each density grid, we obtain the average squared distance between corresponding points of the phase diagrams, i.e. the mean value of the final output measure, such as segregation or inequality, for a given value of parameters in the two setups (isotropic and generated). This average squared distance for each point is then related to the average variance of each of the phase diagram (the reference one and the one for the grid under inquiry). Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability. We tested the sensitivity to the type of distance, using a Minkovski distance with a varying exponent. The results are presented in~\ref{app:distances}, and show a similar sensitivity to spatial initial conditions.

@@

We obtain a very strong sensitivity to initial spatial conditions for the Sugarscape model. Indeed, the relative distance between the phase diagrams of different density grids and the phase diagram of the reference case ranges from 0.09 to 2.98 with a median of 1.52 and an average value of 1.30. The mean distance above 1 means that, on average, the model is more sensitive to the generator parameters than to its own parameters (population and sugar endowment) in the reference model. Moreover, the maximum distance of 2.98 means that the variation due to the change of grid can be up to three times bigger than the variation due to the model parameters. We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in the generator parameter space. Each point represent one of the 50 different density grids used to initialise the distribution of sugar in the model. The points are projected with respect to the generator parameters, and coloured according to the relative distance of the phase diagram of the simulations using this grid to the phase diagram of the reference case. Therefore, the figure ~\ref{fig:sugarscape-distance-meta} shows that the grids generated with a high $\alpha$ (i.e. with a small number of very high density cells) produce simulation results that vary more between the reference case and the generated grid with the same values of parameters than within the reference case because of parameter variations. This pattern is emphasized when grids are generated with a high $\alpha$ and a high $\beta$ (i.e. with low gradient of density decrease around the kernels of high density). These grids have the highest relative distance to the reference case. On the contrary, with grids closer to the uniform pattern of the reference case (bottom left of the graph), the model parameters are more important in determining the final inequality levels than the initial spatial distribution of sugar.

@@

We consider that the method presented in this paper holds great potential for strengthening socio-spatial models' exploration. However, two limits and two areas of opportunities have still not been tackled.

@@

While we investigated the sensitivity to geometry of models together with their indicators, a potential research direction is to study the dependance of indicators themselves to the spatial configuration. % TODO finish

@@

This relates to relations between empirical, conceptual and modeling dimensions of quantitative research in social sciences, as coined by \cite{livet2010ontology}. % TODO finish

@@

After reviewing the extensive literature on spatial biases in statistical and simulation models, we presented a method to analyse the sensitivity of a simulation's results to the initial spatial configuration. We did so by implementing a spatial generator whose output is used as input for the simulation model. We applied this approach to two textbook ABMs: Schelling and Sugarscape. With the Schelling experiment, we found that the different urban morphologies impact the interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index. With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space. We think that this method has the potential to increase the arsenal of evaluation of socio-spatial models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
%%%%%%%%%%%%%
% Acknowledgements

>>>>

\author[5]{Marion Le Texier}
\affil[5]{UMR CNRS 6266 IDEES, Universit{\'e} de Rouen Normandie, France}

@@

\begin{abstract}
Although simulation models of socio-spatial systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually not modified systematically in socio-spatial models, thus leaving unexplored the effect of initial spatial arrangements on the interactions of agents with one another as well as with their environment. In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic geometric structures generator in order to create density grids with which socio-spatial simulation models are initialised. We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models. Furthermore, this effect is sometimes larger than the effect of parameters' value change.
\end{abstract}

@@

Self-organization has been shown to be a central feature of socio-spatial systems in general and of cities in particular ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}. In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, batty2007cities}. Complexity is partially due to bifurcations, which are determinant in socio-spatial systems~\citep{Wilson1981, Wilson2002}. Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures. In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as any other initial conditions, but of more interest to the geographer. \\

@@

In this paper, we aim to provide an operational framework for studying the influence of geometric structures of space on the results of a simulation model representing a sociospatial system. Following \citep{Jessop2008} {\bf Sociospatial systems} are understood here as groups of social agents whose behaviour is constrained by their position in geographical space, in the empirical world. The aim of the geosimulation models we are interested in is to represent, simulate and explain the dynamics of these systems. In the sociospatial systems as well as in the models representing them, agents are in {\bf interaction} among themselves and with their geographical environment. The {\bf geometric structure} of the environment defined at the initialisation of the model and its influence on the model outputs is our object in this paper. We present how to generate the geometric outputs which are used as initial spatial conditions in the model. To this end, we use the example of a density grid generator and feed its outputs to two application models (Schelling and Sugarscape). In no way do we pretend to provide a full exploration of these two particular models, of their attractors and/or potential policy implications. Instead, we present a way of performing a sensitivity analysis to initial spatial conditions of models generated systematically. The generator being controlled by its own parameters, we can then relate the parameters used to generate initial spatial conditions to the variation simulation outcomes. The purpose is two-fold: (i) to test the robustness of simulation results to small variations of generator parameters and (ii) to study the non-trivial effects of typical categories of geometric structures (monocentric \textit{vs.} polycentric for example) on the results of a given model. Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies to our knowledge. In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.

@@

The spatial generator applies an urban morphogenesis model \citep{batty2007cities} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}. An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily into our workflow. Grids are generated through an iterative process which, starting from an empty grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to a patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (i.e. its 8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $t_f$ is reached. The grid then has a population of $t_f \cdot N$ (the population lost due to diffusion process to the outside is reallocated through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four generator parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment. We furthermore fix the total population at $t_f\cdot N = 100,000$, and determine therein the number of steps needed at a given $N$. Typical value ranges for the  parameters will be taken as, following \citet{raimbault2018calibration}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$. We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configurations that can be generated.

@@

The implementations of the models were done from NetLogo. We modified the Sugarscape version with wealth of NetLogo model library (to be able to explore it intensively) and we implemented from scratch the Schelling model. Both pseudo-codes are available in~\ref{app:code}, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}. Density grids are also available at this address. Simulation data are available for reproducibility on the dataverse repository at \url{https://doi.org/10.7910/DVN/A0N5GV}.

@@

We measure the distance of the phase diagrams for all density grids with respect to the reference phase diagram computed on the default initial spatial condition setup (a bi-centric symmetrical non toroidal configuration) using the measure defined in equation~\ref{eq:phase-distance}. For each density grid, we obtain the average squared distance between corresponding points of the phase diagrams, i.e. the mean value of the final output measure, such as segregation or inequality, for a given value of parameters in the two setups (isotropic and generated). This average squared distance for each point is then related to the average variance of each of the phase diagram (the reference one and the one for the grid under inquiry). Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability. We tested the sensitivity to the type of distance, using a Minkovski distance with a varying exponent. The results are presented in~\ref{app:distances}, and show a similar sensitivity to geometric structures.

@@

We obtain a very strong sensitivity to geometric structures for the Sugarscape model. Indeed, the relative distance between the phase diagrams of different density grids and the phase diagram of the reference case ranges from 0.09 to 2.98 with a median of 1.52 and an average value of 1.30. The mean distance above 1 means that, on average, the model is more sensitive to the generator parameters than to its own parameters (population and sugar endowment) in the reference model. Moreover, the maximum distance of 2.98 means that the variation due to the change of grid can be up to three times bigger than the variation due to the model parameters. We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in the generator parameter space. Each point represent one of the 50 different density grids used to initialise the distribution of sugar in the model. The points are projected with respect to the generator parameters, and coloured according to the relative distance of the phase diagram of the simulations using this grid to the phase diagram of the reference case. Therefore, the figure ~\ref{fig:sugarscape-distance-meta} shows that the grids generated with a high $\alpha$ (i.e. with a small number of very high density cells) produce simulation results that vary more between the reference case and the generated grid with the same values of parameters than within the reference case because of parameter variations. This pattern is emphasized when grids are generated with a high $\alpha$ and a high $\beta$ (i.e. with low gradient of density decrease around the kernels of high density). These grids have the highest relative distance to the reference case. On the contrary, with grids closer to the uniform pattern of the reference case (bottom left of the graph), the model parameters are more important in determining the final inequality levels than the initial spatial distribution of sugar.

@@

We consider that the method presented in this paper holds great potential for strengthening socio-spatial models' exploration. However, three limits and two areas of opportunities have still not been tackled.

@@

While we investigated the sensitivity to geometry of models together with their indicators, a potential research direction is to study the dependence of indicators themselves to the spatial configuration. Indeed, the segregation indices are likely to be zoning, size or population density dependant \citep{Wong1997,ReardonOSullivan2004}. Generating point data rather than an grid could help overcoming these issues, and appears as a promising direction for future research. This methodological claim is also interesting from theoretical and empirical views as point data may be used to construct \textit{egohoods} \citep{Hippetal2013}, i.e. a smooth definition of neighbourhoods as opposed to the commonly used non-overlapping units.

@@

This relates to relations between empirical, conceptual and modeling dimensions of quantitative research in social sciences, as coined by \cite{livet2010ontology}. % TODO finish
% to inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, we added a paragraph in conclusion, discussing how our approach allows to go back to questioning empirical cases from the model and questioning our concepts (for example how ’neighbourhood’ is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences)

@@

After reviewing the extensive literature on spatial biases in statistical and simulation models, we presented a method to analyse the sensitivity of a simulation's results to the initial spatial configuration. We did so by implementing a spatial generator whose output is used as input for the simulation model. We applied this approach to two textbook ABMs: Schelling and Sugarscape. With the Schelling experiment, we found that the different urban morphologies impact the interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index. With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space. We think that this method has the potential to increase the arsenal of evaluation of socio-spatial models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.

%%%%%%%%%%%%%
% Acknowledgements
